Symbolic Weighted Language Models and Parsing over Infinite Alphabets

### 2021-05-12 main.tex
intro (fin)

\begin{abstract}
%Symbolic Weighted (SW) extension of symbolic automata...
We propose a framework for weighted parsing over infinite alphabets.

It is based on language models called Symbolic Weighted Automata (SWA) 
at the joint %intersection 
between Symbolic Automata (SA) and Weighted Automata (WA), 
as well as Transducers (SWT) and Visibly Pushdown (SWPDA) variants.
%
Like SA, SWA deal with large or infinite input alphabets, 
and like WA, they output a weight value in a semiring domain.
The transitions of WA are labeled by functions from an infinite alphabet into the weight domain. 
This is unlike WA whose transitions are guarded by boolean predicates 
overs symbols in an infinite alphabet 
and also unlike WA whose transitions are labeled by constant weight values, 
and who deal only with finite automata.
%
We present some properties of WA, WT and SWPDA models, 
%and show how they can be used 
that we use to define and solve a variant of parsing 
over infinite alphabets.
%
We also briefly describe the application that motivated the introduction of these models: 
an parse-based approach to automated music transcription.
\end{abstract}

@@

Other extensions of language models  %(automata and grammars) 
help in dealing with non-determinism, by the computation of weight values. 
With an ambiguous grammar, there may exists several derivations 
(\emph{abstract syntax trees} -- AST), % \ie the result of an analyze.
yielding one input word. % structuring a word, 
The association of one weight value %in semiring domain
to each derivation permits to select a best one (or $n$ bests). % a fixed number of bests 
% = ...a ranking of derivations 
This is roughly the principle of \emph{weighted parsing}
approaches~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}.
In \emph{weighted language models}, 
like \eg probabilistic context-free grammars, % (CFG),
a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with a associative product operator~$\otimes$ into 
the weight of a derivation.
A second operator~$\oplus$, associative and commutative, 
is moreover used to handle ambiguity of the model, 
by summing the weights of the possibly several (in general exponentially many) AST %syntax trees 
associated to a given input word.
Typically, $\oplus$ can select the minimum of two weights.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is assumed, at minima, 
to form a \emph{semiring} $\Semiring$ handling infinite sums, 
%\cite{Eilenberg74automata}
such as a min-plus algebra of domain $\mathbb{R}^+ \cup \{ +\infty\}$, 
where $\oplus$ is min and $\otimes$ is plus.
%by ranking 
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context 
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on 
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful 
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present a uniform framework for weighted parsing over infinite input alphabets.
It is based on some weighted %finite states 
language models generalizing 
both SA, with functions in an arbitrary semiring instead of Boolean guards, 
and Weighted Automata (\textsf{WA}), by handling infinite alphabets -- Figure~\ref{fig:hierarchy}.
%***
%some weighted language models
%model of weighted CFG 
%computing on words over infinite input alphabets %sets of terminal symbol
%- but with finite sets of states and transitions rules. 
% non-terminals and production rules).
In their transition rules, input symbols appear as variables %(parameters)
and the weight associated to a transition rule is a function of these variables.
%
\begin{figure}
\centering
\begin{tikzpicture}
\node (NFA) at (0,4) {%
  \(
  \begin{array}{c} 
  \mathsf{NFA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{B}\\
  q \xrightarrow{a} q' \quad  a \in \Sigma_\mathsf{fin}
  \end{array} 
  \)
};
%
\node (WA) at (-2,2) {%
  \(
  \begin{array}{c} 
  \mathsf{WA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{S}\\
  q \xrightarrow{a, w} q' \quad  a \in \Sigma_\mathsf{fin}, w \in \mathbb{S}
  \end{array} 
  \)
};
%
\node (SA) at (2,2) {%
  \(
  \begin{array}{c} 
  \mathsf{SA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{B}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{B}
  \end{array} 
  \)
};
%
\node (SWA) at (0,0) {%
  \(
  \begin{array}{c} 
  \mathsf{SWA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{S}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{S}
  \end{array} 
  \)
};
\draw[->] (NFA)--(WA);
\draw[->] (NFA)--(SA);
\draw[->] (WA)--(SWA);
\draw[->] (SA)--(SWA);
%\begin{array}{c} \mathsf{NFA} : \Sigma^* \to \mathbb{B} \end{array} 
\end{tikzpicture}
\caption{Classes of Symbolic/Weighted Automata. 
$\Sigma_\mathsf{fin}$ is a finite alphabet, 
$\Sigma_\mathsf{inf}$ is a countable alphabet,
$\mathbb{B}$ is the Boolean algebra, 
$\mathbb{S}$ is an arbitrary commutative semiring, 
$q \xrightarrow{\dots} q'$ represents the form of a transition between states $q$ and $q'$.}
\label{fig:hierarchy}  
\end{figure}
%
%This approach is close to the case of 
%Symbolic Automata (SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM}, 
%except that the domain for weight values is not restricted to be Boolean, 
%like for the guards in the rules of SA, 
%but can be an arbitrary commutative semiring (assuming some restrictions).
%
The models presented here are finite automata called Symbolic Weighted (\SWA),
transducers (\SWT) and pushdown automata,
with a visibly restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
%\wrt an edit distance
The latter model of automata computes on \emph{nested words}~\cite{AlurMadhusudan09nested}, 
a structured form of words parenthesized with markup symbols, 
corresponding to a linearization of trees.
In the context of parsing, there are used here to represent (weighted) AST of CF grammars.
More precisely, a \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$ 
to a given a nested word $t$ which is the linearization of an AST. %representing a parse tree.
%
On the other hand, 
a \SWT is used to define a distance $d(s, t)$ between two finite words $s$ and $t$
over an infinite alphabet, following~\cite{Mohri03EDWA}.
Then, the \emph{SW-parsing} problem aims at %computing 
finding $t$ minimizing (\wrt the ranking defined by $\oplus$)
$d(s, t) \otimes A(w)$ -- 
this value is called the distance between $s$ and $A$ in~\cite{Mohri03EDWA}.

... like above weighted-parsing methods, 
our approach to parsing proceeds in 2 steps, 
Bar-Hillel construction... and best path search \cite{Mohri,Huang05}...
** revise **
%In expressiveness, they are equivalent to weighted CFG. 
%and can be used in a general approach for parsing over infinite input alphabets.
%
%Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$ 
%to a given a nested word $w$ (representing a parse tree),
%and let a \SWT compute a distance $d$, in $\Semiring$, 
%between 2 strings over respectively an infinite input alphabet and the 
%same (infinite) alphabet of $A$.
%Then, the problem of Symbolic Weighted Parsing is, 
%given an input string $s$, to find a nested word $w$ minimizing 
%(according to the ranking defined by $\oplus$)
%the distance $d(s, w) \otimes A(w)$ between $s$ and $A$, 
%as defined in~\cite{Mohri03EDWA}.

@@

The main contributions of the paper are: ...
%generalization of 2 classes
%framework for parsing over infinite alphabet, based on generic def. of distances
%use of VPA for parsing, word instead of trees, use comparison input-output

In Section~\ref{section:SWA} we introduce \SWA and \SWT.
Then \SWVPA are defined in Section~\ref{sec:SWVPA}, 
where a polynomial 1-best algorithm is described that can be use to solve
Symbolic Weighted Parsing, as described Section~\ref{sec:parsing}.
Finally, in Section~\ref{sec:transcription}, we present an application 
of this approach to automated music transcription that has been implemented.

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{|c|c|C{2em}|C{2em}|C{2em}|C{2em}|}
\hline
        & domain & $\oplus$ & $\otimes$ & $\zero$  & $\one$\\ %& nat. ordering\\
\hline\hline
Boolean & $\{\bot, \top\}$ & $\vee$ & $\wedge$ & $\bot$ & $\top$\\ %& $\top \leq_\oplus \bot$  \\
\hline
Counting & $\mathbb{N}$ & $+$ & $\times$ & 0 & 1 \\
\hline
Viterbi & $[0, 1] \subset \mathbb{R}$ & $\mathit{max}$ & $\times$ & 0 & 1\\ % & $x \leq_\oplus y \iff x \ge y$  \\
\hline
Tropical & $\mathbb{R}_+ \cup \{ +\infty\}$ & $\mathit{min}$ & $+$ & $+\infty$ & $0$\\ % & $x \leq_\oplus y \iff x \leq y$   \\
\hline
%MaxPlus & $\mathbb{Q} \cup \{ -\infty\}$ & $\mathsf{max}$ & $+$ & $-\infty$ & $0$ \\
%\hline
%Word lang. & $2^{\Sigma^*}$ & $\cup$ & $\cdot$ & $\emptyset$ & $\{ \varepsilon \}$ \\
%\hline
\end{tabular}
\end{center}
\vskip-1em
\caption{Some commutative, bounded, total and complete semirings.}
\label{fig:semirings}
\end{figure}

@@

\subsection{Best-first Search}
Let us assume that the semiring $\Semiring$ is
commutative, bounded, complete, and total.
% total ? 
% monotonic and superior writ natural ordering
%Regarding the infinite sum operator, note that
%$\bigoplus_{x \in \Phi_\Omega} \phi(x)$, 
%$\bigoplus_{y \in \Phi_\Delta} \psi(y)$, and  
%... exist and in $\Semiring$.
We propose a Dijkstra algorithm computing the minimal weight by $A$, 
\wrt $\leq_\oplus$, for a word in $\Semiring^*$.

@@

\subsection{Trees and Nested-Words}
The hierarchical structure of nested-words, defined the \emph{call} and \emph{return} markup symbols of 
suggest a correspondence between these word and trees. 
The lifting of this correspondence to languages, respective of tree automata and VPA have
been discussed in~\cite{AlurMadhusudan09nested}, 
see also~\cite{Caralp12VPAmult} for the weighted case.
In this section, we discuss the correspondence between the symbolic-weighted extensions
of tree automata and VPA.

>>>>

\begin{abstract}
%Symbolic Weighted (SW) extension of symbolic automata...
We propose a framework for weighted parsing over infinite alphabets.
%
It is based on language models called Symbolic Weighted Automata (SWA) 
at the joint %intersection 
between Symbolic Automata (SA) and Weighted Automata (WA), 
as well as Transducers (SWT) and Visibly Pushdown (SWPDA) variants.
%
Like SA, SWA deal with large or infinite input alphabets, 
and like WA, they output a weight value in a semiring domain.
The transitions of WA are labeled by functions from an infinite alphabet into the weight domain. 
This is unlike WA whose transitions are guarded by boolean predicates 
overs symbols in an infinite alphabet 
and also unlike WA whose transitions are labeled by constant weight values, 
and who deal only with finite automata.
%
We present some properties of WA, WT and SWPDA models, 
%and show how they can be used 
that we use to define and solve a variant of parsing 
over infinite alphabets.
%
We also briefly describe the application that motivated the introduction of these models: 
an parse-based approach to automated music transcription.
\end{abstract}

@@

Other extensions of language models  %(automata and grammars) 
help in dealing with non-determinism, by the computation of weight values. 
With an ambiguous grammar, there may exists several derivations 
(\emph{abstract syntax trees} -- AST), % \ie the result of an analyze.
yielding one input word. % structuring a word, 
The association of one weight value %in semiring domain
to each derivation permits to select a best one (or $n$ bests). % a fixed number of bests 
% = ...a ranking of derivations 
This is roughly the principle of \emph{weighted parsing}
approaches~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}.
In \emph{weighted language models}, 
like \eg probabilistic context-free grammars, % (CFG),
a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with a associative product operator~$\otimes$ into 
the weight of a derivation.
A second operator~$\oplus$, associative and commutative, 
is moreover used to handle ambiguity of the model, 
by summing the weights of the possibly several (in general exponentially many) AST %syntax trees 
associated to a given input word.
Typically, $\oplus$ can select the minimum of two weights.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is assumed, at minima, 
to form a \emph{semiring} $\Semiring$ handling infinite sums, 
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra, see Figure~\ref{fig:semirings}. 
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$, 
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking 
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context 
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on 
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful 
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present a uniform framework for weighted parsing over infinite input alphabets.
It is based on some weighted %finite states 
language models generalizing 
both SA, with functions in an arbitrary semiring instead of Boolean guards, 
and Weighted Automata (\textsf{WA}), by handling infinite alphabets -- Figure~\ref{fig:hierarchy}.
%***
%some weighted language models
%model of weighted CFG 
%computing on words over infinite input alphabets %sets of terminal symbol
%- but with finite sets of states and transitions rules. 
% non-terminals and production rules).
In their transition rules, input symbols appear as variables %(parameters)
and the weight associated to a transition rule is a function of these variables.
%
\begin{figure}
\centering
\begin{tikzpicture}
\node (NFA) at (0,4) {%
  \(
  \begin{array}{c} 
  \mathsf{NFA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{B}\\
  q \xrightarrow{a} q' \quad  a \in \Sigma_\mathsf{fin}
  \end{array} 
  \)
};
%
\node (WA) at (-2,2) {%
  \(
  \begin{array}{c} 
  \mathsf{WA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{S}\\
  q \xrightarrow{a, w} q' \quad  a \in \Sigma_\mathsf{fin}, w \in \mathbb{S}
  \end{array} 
  \)
};
%
\node (SA) at (2,2) {%
  \(
  \begin{array}{c} 
  \mathsf{SA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{B}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{B}
  \end{array} 
  \)
};
%
\node (SWA) at (0,0) {%
  \(
  \begin{array}{c} 
  \mathsf{SWA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{S}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{S}
  \end{array} 
  \)
};
\draw[->] (NFA)--(WA);
\draw[->] (NFA)--(SA);
\draw[->] (WA)--(SWA);
\draw[->] (SA)--(SWA);
%\begin{array}{c} \mathsf{NFA} : \Sigma^* \to \mathbb{B} \end{array} 
\end{tikzpicture}
\caption{Classes of Symbolic/Weighted Automata. 
$\Sigma_\mathsf{fin}$ is a finite alphabet, 
$\Sigma_\mathsf{inf}$ is a countable alphabet,
$\mathbb{B}$ is the Boolean algebra, 
$\mathbb{S}$ is an arbitrary commutative semiring, 
$q \xrightarrow{\dots} q'$ represents the form of a transition between states $q$ and $q'$.}
\label{fig:hierarchy}  
\end{figure}
%
%This approach is close to the case of 
%Symbolic Automata (SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM}, 
%except that the domain for weight values is not restricted to be Boolean, 
%like for the guards in the rules of SA, 
%but can be an arbitrary commutative semiring (assuming some restrictions).
%
The models presented here are finite automata called Symbolic Weighted (\SWA),
transducers (\SWT) and pushdown automata,
with a visibly restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
%\wrt an edit distance
The latter model of automata computes on \emph{nested words}~\cite{AlurMadhusudan09nested}, 
a structured form of words parenthesized with markup symbols, 
corresponding to a linearization of trees.
In the context of parsing, there are used here to represent (weighted) AST of CF grammars.
More precisely, a \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$ 
to a given a nested word $t$ which is the linearization of an AST. %representing a parse tree.
%
On the other hand, 
a \SWT is used to define a distance $T(s, t)$ between two finite words $s$ and $t$
over an infinite alphabet, following~\cite{Mohri03EDWA}.
Then, the \emph{SW-parsing} problem aims at %computing 
finding $t$ minimizing 
$T(s, t) \otimes A(t)$ (\wrt the ranking defined by $\oplus$)
-- this value is called the distance between $s$ and $A$ in~\cite{Mohri03EDWA}.
%
Similarly to weighted-parsing 
methods~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}, 
our approach proceeds in two steps, 
based on properties of the SW models. 
The first step is a Bar-Hillel construction where, 
given a \SWT $T$, a \SWVPA $A$, and an input word $s$, 
a \SWVPA $A_{T, s}$ is built, such that for all $t$, $A_{T, s}(t) = T(s, t) \otimes A(t)$.
In the second step, a best AST $t$ is found by applying best search
similar to shortest distance in graphs~\cite{Mohri02semiring,Huang05kbest} to $A_{T, s}$.
%In expressiveness, they are equivalent to weighted CFG. 
%and can be used in a general approach for parsing over infinite input alphabets.
%
%Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$ 
%to a given a nested word $w$ (representing a parse tree),
%and let a \SWT compute a distance $d$, in $\Semiring$, 
%between 2 strings over respectively an infinite input alphabet and the 
%same (infinite) alphabet of $A$.
%Then, the problem of Symbolic Weighted Parsing is, 
%given an input string $s$, to find a nested word $w$ minimizing 
%(according to the ranking defined by $\oplus$)
%the distance $d(s, w) \otimes A(w)$ between $s$ and $A$, 
%as defined in~\cite{Mohri03EDWA}.

@@

The main contributions of the paper are: 
($i$)~the introduction of automata, \SWA, transducers, \SWT (Section~\ref{section:SWA}),
and visibly pushdown automata \SWVPA (Section~\ref{sec:SWVPA}),
generalizing the corresponding classes of symbolic and weighted models, 
%for the weighted computation on (nested) words over infinite alphabets;  
($ii$)~a polynomial best-search algorithm for \SWVPA, %(Section~\ref{sec:best})
% a framework for parsing over infinite alphabets, 
and ($iii$)~a uniform framework (Section~\ref{sec:parsing}) for parsing over infinite alphabets, 
the keys to which are 
($iii.a$)~the \SWT-based definition of generic edit distances between input and output words,
and ($iii.b$)~the use of nested words, and \SWVPA, 
instead of syntax trees and grammars. %(\S~\ref{sec:trees}).
%
Finally, Section~\ref{sec:transcription} describes
the implemented application %of this approach 
to automated music transcription
that motivated this study.

%based these models and on generic def. of distances;
%the use of VPA for parsing, word instead of trees, use comparison input-output.
%and best-search

%In Section~\ref{section:SWA} we introduce \SWA and \SWT.
%Then \SWVPA are defined in Section~\ref{sec:SWVPA}, 
%where a polynomial 1-best algorithm is described that can be use to solve
%Symbolic Weighted Parsing, as described Section~\ref{sec:parsing}.
%Finally, in Section~\ref{sec:transcription}, we present an application 
%of this approach to automated music transcription that has been implemented.

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{|c|c|C{2em}|C{2em}|C{2em}|C{2em}|}
\hline
        & domain & $\oplus$ & $\otimes$ & $\zero$  & $\one$\\ %& nat. ordering\\
\hline\hline
Boolean & $\{\bot, \top\}$ & $\vee$ & $\wedge$ & $\bot$ & $\top$\\ %& $\top \leq_\oplus \bot$  \\
\hline
Counting & $\mathbb{N}$ & $+$ & $\times$ & 0 & 1 \\
\hline
Viterbi & $[0, 1] \subset \mathbb{R}$ & $\mathit{max}$ & $\times$ & 0 & 1\\ % & $x \leq_\oplus y \iff x \ge y$  \\
\hline
Tropical min-plus & $\mathbb{R}_+ \cup \{ +\infty\}$ & $\mathit{min}$ & $+$ & $+\infty$ & $0$\\ % & $x \leq_\oplus y \iff x \leq y$   \\
\hline
%MaxPlus & $\mathbb{Q} \cup \{ -\infty\}$ & $\mathsf{max}$ & $+$ & $-\infty$ & $0$ \\
%\hline
%Word lang. & $2^{\Sigma^*}$ & $\cup$ & $\cdot$ & $\emptyset$ & $\{ \varepsilon \}$ \\
%\hline
\end{tabular}
\end{center}
\vskip-1em
\caption{Some commutative, bounded, total and complete semirings.}
\label{fig:semirings}
\end{figure}

@@

\subsection{Best-first Search} 
\label{sec:best}\label{sec:search}
Let us assume that the semiring $\Semiring$ is
commutative, bounded, complete, and total.
% total ? 
% monotonic and superior writ natural ordering
%Regarding the infinite sum operator, note that
%$\bigoplus_{x \in \Phi_\Omega} \phi(x)$, 
%$\bigoplus_{y \in \Phi_\Delta} \psi(y)$, and  
%... exist and in $\Semiring$.
We propose a Dijkstra algorithm computing the minimal weight by $A$, 
\wrt $\leq_\oplus$, for a word in $\Semiring^*$.

@@

\subsection{Trees and Nested-Words}
\label{sec:trees}
The hierarchical structure of nested-words, defined the \emph{call} and \emph{return} markup symbols of 
suggest a correspondence between these word and trees. 
The lifting of this correspondence to languages, respective of tree automata and VPA have
been discussed in~\cite{AlurMadhusudan09nested}, 
see also~\cite{Caralp12VPAmult} for the weighted case.
In this section, we discuss the correspondence between the symbolic-weighted extensions
of tree automata and VPA.
